Using best practices for forecasting biodiversity changes in breeding birds

### 2017-07-13 manuscript/draft.Rmd
Make the manuscript knitable
* Fix reference formatting issues
* Fix location of settings.yml
* Fix typos in in line refs
* Adding missing refs

---
title: Using best practices for forecasting biodiversity changes in breeding birds
author:
- David J. Harris
- Shawn D. Taylor
- Ethan P. White
output:
  pdf_document: default
  html_document: default
---

`r settings = yaml::yaml.load_file("settings.yaml")`

@@

Forecasting the future state of ecological systems is increasingly important for
planning, management, and evaluating how well ecological models capture the key
ecological processes governing a system [@clark2001; @houlahan2017;
@dietz2017]. One area of particular interest for forecasting is biodiversity
since it is important for ecosystem function, central to conservation planning,
and expected to be influenced by anthropogenic global change [@cardinale2012;
@diaz2015; @tilman2017]. Anticipating potential changes in biodiversity is
crucial for large scale management and conservation, and for addressing debates
regarding whether local scale diversity is declining and may continue to decline
in response to anthropogenic influences [@diaz2015; @dornelas2014].

@@

However, despite the emerging interest in forecasting species richness and other
aspects of biodiversity, little is known about how effectively we can anticipate
their dynamics. This is due in part to the long time scales over which many
ecological forecasts are applied and the resulting difficulty in assessing
whether the predicted changes occurred [@dietzinreview]. What we do know comes
from a small number of hindcasting studies, where models are built using data on
species occurrence and richness from the past and evaluated on their ability to
predict contemporary patterns [e.g., @algar2009; @distler2015]. These studies
are a valuable first step, but lack several components that are important for
developing good forecasting models and understanding how accurately these models
can predict the future. These components of good forecasting and evaluation (Box
1) broadly involve: 1) expanding the use of data to include biological and
environmental time-series [@treddenick2016]; 2) accounting for uncertainty in
observations, processes, model choice, and forecast evaluation
[@clark2001, @dietz2017]; and 3) conducting meaningful evaluations of the
forecasts by hindcasting, archiving short-term forecasts, and comparing
forecasts to baselines to determine if the forecasts are more accurate than
assuming the system is basically static [@Perretti2013].

@@

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in @hyndman2016] to represent a array of different time-series modeling
approaches. These models can include an autoregressive component (as in the
"naive" model, but with the possibility of longer-term dependencies in the
underlying process), a moving average component (where the noise can have serial
autocorrelation) and an integration/differencing component (so that the analysis
could be performed on sequential differences of the raw data, accommodating more
complex patterns including trends). The `auto.arima` function chooses whether to
include each of these components, and how many terms to include for each one,
using AICc [@hyndman2016]. Since there is no seasonal component to the BBS
time-series, we did not include a season component in these models. Otherwise we
used the default settings for this function. [[Note: should probably record
somewhere what order model the auto.arima function tends to choose.  Eyeballing
the results, seems to be most similar to the naive model, but would be good to
check.]]

@@

For the error metric that evaluates both models' point estimates and their
uncertainty simultaneously, the JSDM and the nonlinear richness model converge
towards the time-series baselines at the longest time-scales assessed. This
convergence at longer time scales highlights the importance of considering how
forecasting method performance changes with the distance into the future being
predicted (Box 1). The value of this approach, and associated forecast horizons,
has been raised for knowing how far into the future a model can be effectively
used for forecasting [@petchy2015], but our results suggest a broader value to
this approach for considering the potential importance of different models and
processes for making forecasts at different scales. [[Should make sure Petchy et
al. don't make this point]] Our results show that for forecasts 1-3 years in the
future, baselines and time-series based approaches outperform all other methods
using integrative metrics. At these time scales, environmental changes are
relatively small and changes in the biota may lag behind changes in the
environment. However, as the timescale of the forecast increases to a decade the
JSDMs and environmental richness models get closer to the accuracy of the
time-series methods and even surpass some of those methods. As the forecast
distance increases the amount of environmental change is expected to increase
and the system will have more time to respond, potentially leading to shifts
towards improved relative performance of the models incorporating environmental
and ecological information. If this shift with forecast distance in the relative
importance of different processes and models continues, this could lead to JSDMs
and environmental richness models outperforming time-series based approaches at
sufficiently long time scales. These results suggest that assessment of how
forecast performance changes with time lag is not only important for determining
how far in the future to forecast, but also for determining models and processes
are most relevant for making forecasts at different time scales.

@@

Future efforts will also need to begin to address the additional uncertainty
that comes from error in forecasting the environmental conditions themselves. In
this and other hindcasting studies the environmental conditions for the "future"
are known because the data has already been observed. In real forecasts the
environmental conditions themselves have to be forecast and those forecasts have
uncertainty and bias. This will cause ecological forecasts that use
environmental data to be more uncertain than observed in current hindcasting
efforts. It is important to correctly incorporate this uncertainty in the
predictor variables into forecasting models [@clark2001; @dietz2017]. Difficulty
in forecasting future environmental conditions at small scales will present
continued challenges for models incorporating these conditions and this may
result in a continued advantage to simple time-series based approaches.

Another challenge for ecological forecasting is the timescales of typical
forecasts. We evaluated forecasts using yearly timesteps up to a decade into the
future. In contrast, many ecological forecasts are temporally aggregated to 5-30
year timesteps and projected up to a century into the future. These are commonly
beyond the career or even lifespan of the researchers, making reasonable model
assessment impossible. Currently assessments at these long time-scales can
only be made with a small number of opportune datasets. For examples hindcasts
have been used to asses models of species richness [@alger2009, @distler2015]
and distribution [@rappacciuolo2012, @moran-ordonez2017, @araujo2005,
@eskildsen2013] using data aggregated across similar time scales. More
evaluation studies like these are needed but there is a paucity of adequate long
term data available. Short term forecasts will allow for relatively rapid
validation of models by researchers without needing to wait for enough data to
accumulate [@tredennick2016; @dietze2016]. This does not mean that they should
replace large scale long-term forecasts. As discussed above, drivers of species
richness are likely to differ at different temporal scales [@rosenzweig1995,
@white2004, @white2007]. Thus short-term forecasts should be expected to add
value to end users and inform, not replace, long-term forecasts. Hopefully by
evaluating trends in the shorter term forecasts as done here it will be
possible to also use the information from short-term forecasting to improve
longer-term estimates.

The science of forecasting biodiversity, and ecology more broadly, remains in
its infancy and it is important to consider the general inability of
forecasting methods to improve on simple baselines in that context. When weather
forecasting first started the forecasts were likewise worse than simple
baselines [@mcgill2012]. One of the things that helped weather forecasts improve
was large numbers of forecasts were made in public, which allowed different
approaches to forecasting to be regularly assessed and improved [@mcgill2012;
@silver2012]. This suggests that it is important for ecologists to start
regularly making and evaluating true ecological forecasts, even if they perform
poorly, and to make these forecasts publicly available for assessment. These
forecasts should include both short-term predictions, which can be assessed
quickly, and mid to long-term forecasts to help assess long time-scale processes
and determine how far into the future we can successfully forecast
[@tredenick2016; @dietz2017]. Forecasts based on the models in this paper from
now until 2050 are openly archived on Dryad so that we and others can assess how
well they perform and we plan to evaluate these forecasts and report the results
as each new year of BBS data becomes available. Weather forecasting has
continually improved throughout its history [@bauer2015] in part due to making
and evaluating public forecasts [@mcgill2012] and we hope this will help ecology
do the same.

@@

Forecasts describe how systems are expected to change through time. While some
areas of ecological forecasting focus primarily on time-series data, others
primarily focus on using spatial models and space-for time substitutions. Using
ecological and environmental time-series data allows the consideration of actual
dynamics from both a process and error structure perspective [@treddenick2016a].

@@

Understanding uncertainty in a forecast is just as important as understanding
the average or expected outcome. Failing to account for uncertainty can result
in overconfidence in highly uncertain outcomes leading to poor decision making
and erosion of confidence in ecological forecasts. Models should explicitly
include sources of uncertainty and propagate them through the forecast where
possible [@clark2001; @dietz2017]. Evaluations of forecasts should assess the
accuracy of uncertainties as well as point estimates [@dietz2017].

@@

In general the accuracy of forecasts decreases with the length of time into the
future being forecast [@petch2015]. This decay in accuracy and the potential for
different rates of decay to result in different relative model performance at
different lead times should be considered when evaluating forecasts and
comparing models.

@@

In cases where long-term predictions are the primary goal, short-term should
also be made to accommodate the time-scales of planning and management decisions
and to allow the accuracy of the forecasts to be quickly evaluated
[@treddenick2016].

>>>>

---
title: Using best practices for forecasting biodiversity changes in breeding birds
author:
- David J. Harris
- Shawn D. Taylor
- Ethan P. White
bibliography: refs.bib
output:
  pdf_document: default
  html_document: default
---

`r settings = yaml::yaml.load_file("../settings.yaml")`

@@

Forecasting the future state of ecological systems is increasingly important for
planning, management, and evaluating how well ecological models capture the key
ecological processes governing a system [@clark2001; @houlahan2017;
@dietze2017]. One area of particular interest for forecasting is biodiversity
since it is important for ecosystem function, central to conservation planning,
and expected to be influenced by anthropogenic global change [@cardinale2012;
@diaz2015; @tilman2017]. Anticipating potential changes in biodiversity is
crucial for large scale management and conservation, and for addressing debates
regarding whether local scale diversity is declining and may continue to decline
in response to anthropogenic influences [@diaz2015; @dornelas2014].

@@

However, despite the emerging interest in forecasting species richness and other
aspects of biodiversity, little is known about how effectively we can anticipate
their dynamics. This is due in part to the long time scales over which many
ecological forecasts are applied and the resulting difficulty in assessing
whether the predicted changes occurred [@dietze2016]. What we do know comes
from a small number of hindcasting studies, where models are built using data on
species occurrence and richness from the past and evaluated on their ability to
predict contemporary patterns [e.g., @algar2009; @distler2015]. These studies
are a valuable first step, but lack several components that are important for
developing good forecasting models and understanding how accurately these models
can predict the future. These components of good forecasting and evaluation (Box
1) broadly involve: 1) expanding the use of data to include biological and
environmental time-series [@tredennick2016]; 2) accounting for uncertainty in
observations, processes, model choice, and forecast evaluation
[@clark2001, @dietze2017]; and 3) conducting meaningful evaluations of the
forecasts by hindcasting, archiving short-term forecasts, and comparing
forecasts to baselines to determine if the forecasts are more accurate than
assuming the system is basically static [@perretti2013].

@@

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in @forecast1] to represent a array of different time-series modeling
approaches. These models can include an autoregressive component (as in the
"naive" model, but with the possibility of longer-term dependencies in the
underlying process), a moving average component (where the noise can have serial
autocorrelation) and an integration/differencing component (so that the analysis
could be performed on sequential differences of the raw data, accommodating more
complex patterns including trends). The `auto.arima` function chooses whether to
include each of these components, and how many terms to include for each one,
using AICc [@forecast1]. Since there is no seasonal component to the BBS
time-series, we did not include a season component in these models. Otherwise we
used the default settings for this function. [[Note: should probably record
somewhere what order model the auto.arima function tends to choose.  Eyeballing
the results, seems to be most similar to the naive model, but would be good to
check.]]

@@

For the error metric that evaluates both models' point estimates and their
uncertainty simultaneously, the JSDM and the nonlinear richness model converge
towards the time-series baselines at the longest time-scales assessed. This
convergence at longer time scales highlights the importance of considering how
forecasting method performance changes with the distance into the future being
predicted (Box 1). The value of this approach, and associated forecast horizons,
has been raised for knowing how far into the future a model can be effectively
used for forecasting [@petchey2015], but our results suggest a broader value to
this approach for considering the potential importance of different models and
processes for making forecasts at different scales. [[Should make sure Petchey et
al. don't make this point]] Our results show that for forecasts 1-3 years in the
future, baselines and time-series based approaches outperform all other methods
using integrative metrics. At these time scales, environmental changes are
relatively small and changes in the biota may lag behind changes in the
environment. However, as the timescale of the forecast increases to a decade the
JSDMs and environmental richness models get closer to the accuracy of the
time-series methods and even surpass some of those methods. As the forecast
distance increases the amount of environmental change is expected to increase
and the system will have more time to respond, potentially leading to shifts
towards improved relative performance of the models incorporating environmental
and ecological information. If this shift with forecast distance in the relative
importance of different processes and models continues, this could lead to JSDMs
and environmental richness models outperforming time-series based approaches at
sufficiently long time scales. These results suggest that assessment of how
forecast performance changes with time lag is not only important for determining
how far in the future to forecast, but also for determining models and processes
are most relevant for making forecasts at different time scales.

@@

Future efforts will also need to begin to address the additional uncertainty
that comes from error in forecasting the environmental conditions themselves. In
this and other hindcasting studies the environmental conditions for the "future"
are known because the data has already been observed. In real forecasts the
environmental conditions themselves have to be forecast and those forecasts have
uncertainty and bias. This will cause ecological forecasts that use
environmental data to be more uncertain than observed in current hindcasting
efforts. It is important to correctly incorporate this uncertainty in the
predictor variables into forecasting models [@clark2001; @dietze2017]. Difficulty
in forecasting future environmental conditions at small scales will present
continued challenges for models incorporating these conditions and this may
result in a continued advantage to simple time-series based approaches.

Another challenge for ecological forecasting is the timescales of typical
forecasts. We evaluated forecasts using yearly timesteps up to a decade into the
future. In contrast, many ecological forecasts are temporally aggregated to 5-30
year timesteps and projected up to a century into the future. These are commonly
beyond the career or even lifespan of the researchers, making reasonable model
assessment impossible. Currently assessments at these long time-scales can
only be made with a small number of opportune datasets. For examples hindcasts
have been used to assess models of species richness [@algar2009, @distler2015]
and distribution [@rapacciuolo2012, @moran-ordonez2017, @araujo2005,
@eskildsen2013] using data aggregated across similar time scales. More
evaluation studies like these are needed but there is a paucity of adequate long
term data available. Short term forecasts will allow for relatively rapid
validation of models by researchers without needing to wait for enough data to
accumulate [@tredennick2016; @dietze2016]. This does not mean that they should
replace large scale long-term forecasts. As discussed above, drivers of species
richness are likely to differ at different temporal scales [@rosenzweig1995,
@white2004, @white2007]. Thus short-term forecasts should be expected to add
value to end users and inform, not replace, long-term forecasts. Hopefully by
evaluating trends in the shorter term forecasts as done here it will be
possible to also use the information from short-term forecasting to improve
longer-term estimates.

The science of forecasting biodiversity, and ecology more broadly, remains in
its infancy and it is important to consider the general inability of
forecasting methods to improve on simple baselines in that context. When weather
forecasting first started the forecasts were likewise worse than simple
baselines [@mcgill2012]. One of the things that helped weather forecasts improve
was large numbers of forecasts were made in public, which allowed different
approaches to forecasting to be regularly assessed and improved [@mcgill2012;
@silver2012]. This suggests that it is important for ecologists to start
regularly making and evaluating true ecological forecasts, even if they perform
poorly, and to make these forecasts publicly available for assessment. These
forecasts should include both short-term predictions, which can be assessed
quickly, and mid to long-term forecasts to help assess long time-scale processes
and determine how far into the future we can successfully forecast
[@tredennick2016; @dietze2017]. Forecasts based on the models in this paper from
now until 2050 are openly archived on Dryad so that we and others can assess how
well they perform and we plan to evaluate these forecasts and report the results
as each new year of BBS data becomes available. Weather forecasting has
continually improved throughout its history [@bauer2015] in part due to making
and evaluating public forecasts [@mcgill2012] and we hope this will help ecology
do the same.

@@

Forecasts describe how systems are expected to change through time. While some
areas of ecological forecasting focus primarily on time-series data, others
primarily focus on using spatial models and space-for time substitutions. Using
ecological and environmental time-series data allows the consideration of actual
dynamics from both a process and error structure perspective [@tredennick2016].

@@

Understanding uncertainty in a forecast is just as important as understanding
the average or expected outcome. Failing to account for uncertainty can result
in overconfidence in highly uncertain outcomes leading to poor decision making
and erosion of confidence in ecological forecasts. Models should explicitly
include sources of uncertainty and propagate them through the forecast where
possible [@clark2001; @dietze2017]. Evaluations of forecasts should assess the
accuracy of uncertainties as well as point estimates [@dietze2017].

@@

In general the accuracy of forecasts decreases with the length of time into the
future being forecast [@petchey2015]. This decay in accuracy and the potential for
different rates of decay to result in different relative model performance at
different lead times should be considered when evaluating forecasts and
comparing models.

@@

In cases where long-term predictions are the primary goal, short-term should
also be made to accommodate the time-scales of planning and management decisions
and to allow the accuracy of the forecasts to be quickly evaluated
[@tredennick2016].
